---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# gamutil

<!-- badges: start -->
[![R-CMD-check](https://github.com/msaito8623/gamutil/workflows/R-CMD-check/badge.svg)](https://github.com/msaito8623/gamutil/actions)
<!-- badges: end -->

The goal of gamutil is to provide some short-cut functions to facilitate to work with mgcv::gam and mgcv::bam.

## Installation

gamutil is still under development and therefore not available on [CRAN](https://CRAN.R-project.org). 

The development version is available from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("msaito8623/gamutil")
```

## Examples

This is a basic example which shows you how to visualize prediction by a GA(M)M model with a contour plot:

```{r, ex1, fig.dim=c(6,6), out.width="600px", out.height="500px"}
library(gamutil)
library(mgcv)
set.seed(8361)
dat = gamSim(verbose=FALSE)
mdl = gam(y ~ s(x0) + s(x1) + ti(x0,x1), data=dat)
plt = plot_contour(mdl, view=c('x0','x1'))
print(plt)
```

---

The argument "axis.len" controls resolution of the contour plot to be produced (default=10). Bigger numbers for the argument would draw smoother contour lines.


```{r, ex2, fig.dim=c(6,6), out.width="600px", out.height="500px"}
plt = plot_contour(mdl, view=c('x0','x1'), axis.len=100)
print(plt)
```

---

The interval between contour lines can be adjusted by the "break.interval" argument.

```{r, ex3, fig.dim=c(6,6), out.width="600px", out.height="500px"}
plt = plot_contour(mdl, view=c('x0','x1'), axis.len=100, break.interval=1.5)
print(plt)
```
